---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: yes
---


## Loading and preprocessing the data
``` {r Part1}
    data <- read.csv("./activity.csv")
    ## Create subsets of original data for different parts of the assignment
    dt1 <- data[complete.cases(data),]
    get_steps_by_date <- function(x) {
      split(x[,"steps"], x[,"date"])
    }
    get_steps_by_int <- function(x) { 
      split(x[,"steps"], x[,"interval"])
    }


```  

## What is mean total number of steps taken per day?
``` {r Part2}
exec_part2 <- function(x) {
      steps_by_date <- get_steps_by_date(x)
    
      ## Total number of steps taken each day
      total_steps_by_date <- sapply(steps_by_date, sum)
      summary(total_steps_by_date) 
      hist(total_steps_by_date, main = "Histogram of Total Steps per Day", xlab = "Total Steps per Day")
      
      ## Mean of the Total number of steps taken each day
      mean_steps_by_day <- sapply(steps_by_date, mean)
      tot_mean <- mean(mean_steps_by_day, na.rm = T)
      
      ## Median of the Total number of steps taken each day
      median_steps_by_day <- sapply(steps_by_date, median)
      tot_med <- median(median_steps_by_day, na.rm = T)
      
      gettmn <- function() { tot_mean }
      gettmdn <- function() { tot_med }
      
      list(gettmdn = gettmdn, gettmn = gettmn, total_steps_by_date = total_steps_by_date, mean_steps_by_day = mean_steps_by_day, median_steps_by_day = median_steps_by_day)
}

    tot_vals <- exec_part2(dt1)
    tot_mean1 <- tot_vals$gettmn()
    tot_steps1 <- tot_vals$total_steps_by_date
    
    tot_vals$median_steps_by_day
    tot_vals$mean_steps_by_day
```  

The Mean of the Total number of steps taken each day is : **`r tot_vals$gettmn()`**  
The Median of the Total number of steps taken each day is : **`r tot_vals$gettmdn()`**  


## What is the average daily activity pattern?
``` {r Part3}

exec_part3 <- function(x) {
      steps_by_int <- get_steps_by_int(x)
      int <- names(steps_by_int)
      
      ## Mean Steps by Interval
      mean_steps_by_int <- sapply(steps_by_int, mean)
      
      ## Obtain the max of the mean steps per interval and its interval name
      tot_max <- max(mean_steps_by_int, na.rm = T)
      max_int <- names(mean_steps_by_int[which.max(mean_steps_by_int)])
      
      gettmax <- function() { tot_max }
      
      list(gettmax = gettmax, int = int, max_int = max_int, mean_steps_by_int = mean_steps_by_int)
  
}

    tot_vals <- exec_part3(dt1)
    mean_steps_by_int <- tot_vals$mean_steps_by_int
    
    ## Assigning a meaningful name to the Intervals for easier understanding
    intervals <- strptime(sprintf("%04d", as.numeric(names(mean_steps_by_int))), format="%H%M")
    
    ## Plotting the intervals against the Avg number of steps taken per Interval across all days
    plot(intervals, mean_steps_by_int, type = "l", xlab="Interval (hh:mm)", ylab="Steps per interval")
    
    tot_max <- tot_vals$gettmax()
    int <- tot_vals$int
    max_int <- tot_vals$max_int

```  
On Average across all days, the Interval - **`r max_int`** - contains the Maximum number of Steps which is equal to - **`r tot_max`**  

## Imputing missing values  

``` {r Part4}
    ## Count of Missing Values in the dataset  
    cnt_na <- sum(! complete.cases(data))
```
The total number of missing values in the dataset are : **`r cnt_na`**  

``` {r Part5}

    ## Create a data table to store the new imputed dataset
    require(data.table)
    data2 <- as.data.table(data)
    
    ## Fill-in the missing values with the mean of steps per interval for each missing value
    data2[is.na(data2$steps), steps := as.integer(mean_steps_by_int[int])]
    
    ## Check if there are any NAs still left in the new data table
    if (sum(! complete.cases(data2)) == 0) {
      print("All Missing Values are imputed")
    }
    
    tot_vals <- exec_part2(setDF(data2))
    tot_mean2 <- tot_vals$gettmn()
    tot_steps2 <- tot_vals$total_steps_by_date
    
    stps_diff <- abs(tot_steps1 - tot_steps2)
    stps_diff_dts <- names(stps_diff[which(stps_diff != 0)])

```  

The Mean of the Total number of steps taken each day (after imputation) is : **`r tot_vals$gettmn()`**  
The Median of the Total number of steps taken each day (after imputation) is : **`r tot_vals$gettmdn()`** 

The Difference between the values of mean from the first part of assignment (i.e., before imputation and after imputation) is : **`r (tot_mean1 - tot_mean2)`**

There is no difference between the median values before and after the imputation.

The Difference between the values of total steps per day from the first part of assignment (i.e., before imputation and after imputation) is observed for the following dates: 
**`r stps_diff_dts`**

## Are there differences in activity patterns between weekdays and weekends?

``` {r Part6}

    ## Create a Factor Variable containing two levels one each for weekday and weekend
    type_of_day <- factor(c("Monday"="0", "Tuesday"="0", "Wednesday"="0", "Thursday"="0", "Friday"="0", "Saturday"="1", "Sunday"="1"), labels = c("weekday","weekend"))
    
    ## Setting the imputed dataset back to data table
    setDT(data2)
    
    ## Add a new column for loading the type of day
    data2[, type_of_day := type_of_day[weekdays(as.Date(date))]]
    
    ## Add another column for storing the mean grouped by interval and type of day
    mn <- sapply(split(data2$steps, list(data2$interval, data2$type_of_day)), mean)
    data2[, mn := mn]
    
    ## Set the key cols so that unique function can be applied on the data table
    keycols <- c("interval", "type_of_day")
    setkeyv(data2, keycols)
    data2 <- unique(data2, by = key(data2))
    
    ## Extract only the interest columns - interval, type_of_day, mean_steps
    setDF(data2)
    df <- data2[, c(3:5)]
    
    ## Build the Plot for each type of day for the mean steps against the interval
    library(lattice)
    xyplot(mn ~ interval | type_of_day, 
           type="l", 
           data = df, 
           layout=c(1,2),
           ylab = "Number of steps",
           xlab = "Interval")

```

As can be observed above plots, on average across all days, the Maximum number of Steps taken is higher during weekdays than over weekends. 